Reconstructing the last known movements of one of Nature's giants

### 2019-01-17 manuscript/PeerJ/blue-new.tex
Update to author year citations

\usepackage[osf]{mathpazo} % palatino
\usepackage{ms}            % load the template
%\usepackage[round]{natbib} % author-year citations
\usepackage{graphicx}
\usepackage{parskip} 
\usepackage{caption}
\usepackage{subcaption}
\usepackage{textcomp} % for parts per mille symbol     
\pagenumbering{arabic}    
\linespread{1.66}

@@

Migratory species pose a particular challenge for conservation practitioners because their effective conservation relies on protection at many, often distant, sites\cite{runge2014conserving}. 
Migratory species may also be particularly vulnerable to changes in climate or human use of the environment, as they are influenced by conditions in multiple locations across different parts of their life cycle \cite{robinson2009travelling}. 
Identifying threats to migratory species, understanding species responses to change and developing effective conservation measures all require information on the movements of individual animals over multiple years, ideally for both historic and present-day populations. 
With the development of electronic tagging technology, studies of the distributions of animals have largely shifted from infrequent observations of many individuals with limited or no information about individual movement behavior, to frequent observations of a few individuals, with detailed information about individual movement\cite{holdo2013inferring}. 
Despite the tremendous advances made using direct telemetry devices, tags are expensive, with a relatively high failure rate \cite{bailey2009behavioural,best2015tag,mate2007evolution}. 
Data on individual-level, multi-annual movements remain scarce, especially for rare, wide-ranging and long-lived marine species such as baleen whales (Mysticeti)\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.

An alternative technique for investigating animal movements retrospectively is to use intrinsic biochemical information such as stable isotope compositions \cite{west2006stable,busquets2017estimating,hobson2008tracking}. 
Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioral information that are often under-utilised\cite{lister2011natural}. 
The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
A large literature describes the use of stable isotope markers to link the composition of a tissue to a geographic source at a single point in time\cite{hobson2008tracking}. 
Relating the isotopic compositions of marine animal tissues to the likely location of tissue growth is complicated by a lack of knowledge of spatial variations in the isotopic composition of diet (the isotopic baseline)\cite{west2006stable,mcmahon2015millennial}. 
This uncertainty is compounded when multiple samples are taken through time in the same individual, as temporal variation in the isotopic baseline must also be considered explicitly. 
Consequently, relatively little attention has been given to the potential to infer individual movement histories by reconstructing time series of multiple discrete origins from sequentially sampled incrementally-grown tissues.

Here we introduce a new approach to infer location from the isotopic compositions of sequentially-sampled incrementally-grown tissues by drawing on newly developed models predicting spatio-temporal variation in phytoplankton stable carbon isotope composition at global and monthly resolution\cite{magozzi2017using} (Fig. S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1; Fig. S5).
We use our coupled models to simulate time series of carbon stable isotope compositions expected under different movement behaviours. 
We then compare simulated isotopic profiles with the measured profiles to identify potential movement histories most consistent with the measured isotopic profile.

We apply our method to blue whales (\textit{Balaenoptera musculus}), the largest animal to have ever lived. 
Mysticete whales are characterized by the development of baleen, keratinous structures in the upper jaw used to filter food items from seawater. 
Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert\cite{best1996stable,hobson1998stable}. 
Baleen plates therefore offer a continuous isotopic record of behavior typically reflecting multiple years of life of an individual whale. 
Baleen is worn away at the tips over time, so a baleen plate reflects the most recent years of life, and rarely records an individual's entire lifespan. 
Among the mysticete whales, blue whales are a particularly attractive target for model-based isotope movement work as they have a consistent, low trophic level diet and feed continuously through the year, likely driven by high energetic costs of maintaining extreme muscle mass\cite{goldbogen2015}.

Similar to other large balaenopterid whales, blue whales are generally assumed to conduct annual migrations between high and low latitudes\cite{huckgaete2018}. 
As blue whales feed throughout the year to accommodate energetic costs, migration routes and feeding areas for blue whales are thought to be shaped by the year-round location of highly productive regions\cite{branch2007}. 
Recent studies tracking movements of individual blue whales primarily in the northeast Pacific have demonstrated high levels of among-individual variation in movement histories including suspending migration for opportunistic feeding and skipped migration (i.e. year round residency in either summer or winter
areas) \cite{busquets2017estimating}. 
The northeast Pacific population of blue whales is perhaps the best studied in terms of individual-level movements with a recent history of deployment of over 177 satellite tags \cite{irvine2017quantifying, mate2007evolution}.
In the northeast Pacific, satellite tags show individuals occupying most of the exclusive economic zone (EEZ) and into the Gulf of Alaska \cite{irvine2017quantifying} with greatest overlapping areas in the southern half of the EEZ, along the continental slope and shelf region between the California Channel Islands and Cape Mendocino. 
Winter feeding is less well characterized, but winter feeding grounds are known
in the Gulf of California and in the eastern tropical Pacific, specifically the Costa Rica Dome where production is maintained by seasonal upwelling.
However, most satellite deployments on blue whales report movement data for periods of time under six months \cite{heide2001new, silva2013north, bailey2009behavioural, lesage2017foraging, irvine2017quantifying}, and are unable to show individual movements between summer and winter feeding or breeding areas. 
The longest period of continuous monitoring via satellite tagging of an individual blue whale that we are aware of in the northeast Pacific is 504 days \cite{irvine2017quantifying} and in the North Atlantic is 177 days \cite{lesage2017foraging}.

The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now around 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland \cite{pike2009note}. 
The small population size and oceanic habit makes blue whale movements extremely hard to study. 
A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggests that blue whales in the northeast Atlantic also track regions of high production throughout the year, at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde Islands \cite{baines2014upwellings}. 
Northward migration in the spring may occur along mid Atlantic corridors with peak sightings in the Azores around April-May \cite{silva2013north}. 
Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas \cite{pike2009note}. 
Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southerly migrating animals \cite{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
To our knowledge one photographic identification has matched summer and winter locations of a single north-east Atlantic blue whale with sightings in Iceland and Mauritania \cite{poster}.

@@

Here we infer individual level high-resolution movement records for seven blue whales for which baleen stable isotope data were available.
Isotope data for six whales stranding on the west coast of the USA or Mexico were reported in Busquets-Vass \textit{et al}.\cite{busquets2017estimating}, and we report new isotopic data from a whale stranding off the coast of Wexford Ireland in March 1891, and currently on display in the Natural History Museum, London (NHM).
The NHM whale ``Hope'' is a female estimated to be a young adult at least 15 years at death. 
Baleen from the NHM whale therefore yields information about migratory behaviors in North Atlantic blue whales around the peak of industrial whaling.

@@

Baleen was collected from the Natural History Museum, London (specimen NHMUK.1892.3.1.1). 
The baleen plate was cleaned with ethanol to remove surface contaminants such as skin/gum or other lipids that can influence isotopic signals. 
{\raise.17ex\hbox{$\scriptstyle\sim$}}1mg samples of keratin powder were then collected from the plate using a hand-held drill and grinding bit. 
97 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal (gingival) section that contains the most recent tissue. 
Baleen grows at a constant rate, so the samples are equally spaced through time\cite{best1996stable}. 
Carbon and nitrogen isotope analysis was performed simultaneously via continuous-flow isotope ratio mass spectrometry at the University of Southampton SEAPORT Stable Isotope Ratio Mass Spectrometry Laboratory (Southampton, UK), using a Vario Isotope select elemental analyser, coupled to an Isoprime 100 isotope mass spectrometer. 
Replicates using internal laboratory standards (L-glutamic acid (C), Glutamic acid (CT standard), acetanilide and protein standard OAS) were used for quality control and calibration. 
C:N ratios for samples ranged from 3.28\text{\textperthousand} to 3.72\text{\textperthousand}, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples\cite{hobson1998stable}. 
All data are available from the NHM Data Portal\cite{data-set} (https://doi.org/10.5519/0093278).

@@

Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates\cite{hobson1998stable,busquets2017estimating}. 
Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the youngest 70 - 18cm of the plate (behavioural phase two). 
We calculated isotopic periodicity within the NHM baleen sample using Fourier Transform analysis\cite{cardona2017temporal} (Supplementary Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated by Busquets-Vass \textit{et al.}\cite{busquets2017estimating}.  
Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891.

@@

Isotope-enabled biogeochemical ocean models \cite{magozzi2017using,schmittner2016complementary} were used to characterize
the isotopic composition of phytoplankton expected in different
potential foraging grounds (Fig. S3). 
Annual average \(\delta^{15}\)N POM (particulate organic matter) values were provided by C.J. Somes (\textit{pers.comm}) based on a 5\({}^{\circ}\) resolution biogeochemical model (Fig. S3). \(\delta^{13}\)C POM values were simulated at 1\({}^{\circ}\) and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model \cite{magozzi2017using,yool2013medusa}. 
Simulated \(\delta^{15}\)N POM values are relatively positive in the northeast Atlantic north of c. 60\({}^{\circ}\)N, and relatively negative in the central and southern North Atlantic. 
Annual average \(\delta^{13}\)C POM values largely vary with latitude, with more negative values in more northerly regions. 
In the central North Atlantic, \(\delta^{13}\)C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. S3). 
The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature \cite{magozzi2017using,laws1995dependence}. 
Thus temporal variations in \(\delta^{13}\)C POM values are superimposed on latitudinal gradients. 
The scale and nature of temporal variation in \(\delta^{13}\)C POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in \(\delta^{13}\)C POM values linked to strongly seasonal phytoplankton growth dynamics.
The model simulations used are forced with decadal climatological data from 2000-2010\cite{magozzi2017using}. 
We do not draw inferences about location from absolute \(\delta^{13}\)C values and thus the reduction in oceanic \(\delta^{13}\)C POM values caused by release of fossil carbon into the atmosphere (Suess effect) does not influence inferences about location or movement. 
Ocean basin scale spatio-temporal differences in \(\delta^{13}\)C values largely reflect relative differences in sea surface temperatures between trophic, temperate and arctic waters, therefore we are confident that the Magozzi \textit{et al}.\cite{magozzi2017using} model can be applied to interpret historic baleen isotope data.

We used \(\delta^{13}\)C POM values modeled at monthly resolution to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours. 
The stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on in the weeks prior to keratin growth. 
Assimilation of carbon into krill tissues will dampen the temporal variability seen in POM, effectively producing a temporal average over the timescale of isotopic turnover within krill. 
We estimate turnover to be complete between two and four months and therefore we resampled the \(\delta^{13}\)C POM values in each one degree cell to reflect an average of isotopic compositions in phytoplankton in the two months prior to the sampling date. 
Average values were weighted according to the proportional plankton biomass estimated for each month. 
Carbon isotope values are also likely to be fractionated during transfer from plankton to krill as \textsuperscript{12}C is preferentially lost through respiration. 
The degree of such trophic fractionation is unclear, however and we do not draw interpretations based on absolute \(\delta^{13}\)C values, rather on the relative \(\delta^{13}\)C values across the length of the baleen plate. 
We assume opportunistic ingestion of protein throughout the year \cite{baines2017autumn,silva2013north,visser2011timing,busquets2017estimating, lesage2017foraging, bailey2009behavioural, branch2007, huckgaete2018}.

@@

We simulated whale movements with the likelihood, direction, and extent of movement influenced by behavioral state, sea surface temperature, water depth, and phytoplankton concentration (as a proxy for zooplankton food availability).
Movement was coded as a set of probabilistic rules, informed by the literature on blue whale behavior (e.g. \cite{handbook}).
All terms were expressed as probability distributions, yielding multiple potential movement tracks.

In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and linear distance of movement are all influenced by the following. 
(i) Behavioural state (migrating north, migrating south, or foraging).
Northerly migrations were possible only in spring, and southerly migrations in autumn. 
Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton.
(ii) Sea surface temperature \cite{yool2013medusa} ($^{\circ}$C). 
When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters.  
(iii) Water depth (m; from the General Bathymetric Chart of the Oceans (GEBCO) global bathymetry dataset\cite{bathy}, extracted via the marmap package in R \cite{marmap}). 
Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
(iv) Phytoplankton concentration ($mmolNm^{-3}$, for combined diatom and non-diatom communities \cite{yool2013medusa}). 
This was included as a proxy for zooplankton food availability. 
Whales are more likely to move towards (or remain within) areas of high phytoplankton density, particularly during the foraging behavioural state.

@@

For the historic North Atlantic whale, we simulated the isotopicexpression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area \cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplemental Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
Measured tracks can only be replicated by combination of residency and seasonal migration with latitudinal migrations limited to the last four
years of simulations. 
For a full description of the model and its parameters see Supplementary Methods: Fig. S5 and Table S1. 
We simulated seven years of whale movements 1000 times and two years of residency 30 times for each residency hotspot (Fig. S4).
We simulated whale movements 1200 times, then excluded simulations where the virtual whale stranded before reaching the 3019 days of the baleen record, leaving 1049 simulations.

At each time point and location in the simulations, we extracted time averaged \(\delta^{13}\)C values from the spatio-temporal model\cite{magozzi2017using}resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figs X, \ref{fig2}, and S4).
We then compared the simulated stable isotope profiles (Figure \ref{fig3}) to the profile of the blue whale (Figure \ref{fig1}).

R code for all analyses is available from GitHub (https://github.com/nhcooper123/blue-whale-bes; \cite{github})

@@

Estimates of baleen growth rates for the six modern whales from the northeast Pacific were inferred from cyclical fluctuations in nitrogen isotope ratios and reported in Busquets-Vass \textit{et al}\cite{busquets2017estimating}. 
We used these growth rates to reconstruct a potential timeline for baleen isotope records.

@@

In four of the six cases, measured baleen carbon isotope profiles could be replicated from model simulations with relatively high precision (Fig. \ref{pacific}
; best fit r\textsuperscript{2} values ranging from 0.43 to 0.74).
Whales D and F showed similar carbon isotope records and were best fit by the same movement simulation. 
We extracted monthly locations associated with the best fitting simulation models for each whale (Fig. \ref{pacific}).
Best fitting simulation models for whales D and F imply a relatively southerly distribution with few locations north of Baja, California and relatively limited seasonal latitudinal migration. 
The measured carbon isotope record for whale A was best fitted by simulations with a southerly distribution, but with an additional strong seasonal latitudinal migration, and with more northerly locations between April and October. 
Whale C was best fit by models with relatively northerly locations and strong seasonal latitudinal migration. 
We were unable to find convincing fits to baleen profiles for whales B and E suggesting that longer simulations with a greater variation in movements were needed. 
The inferred movement records are consistent with the known spatial distributions and relatively high level of individual variation in spatial distributions among blue whales in the Pacific northwest \cite{irvine2017quantifying, mate2007evolution}.

>>>>

\usepackage[osf]{mathpazo} % palatino
\usepackage{ms}            % load the template
\usepackage[round]{natbib} % author-year citations
\usepackage{graphicx}
\usepackage{parskip} 
\usepackage{caption}
\usepackage{subcaption}
\usepackage{textcomp} % for parts per mille symbol     
\pagenumbering{arabic}    
\linespread{1.66}

@@

Migratory species pose a particular challenge for conservation practitioners because their effective conservation relies on protection at many, often distant, sites \citep{runge2014conserving}. 
Migratory species may also be particularly vulnerable to changes in climate or human use of the environment, as they are influenced by conditions in multiple locations across different parts of their life cycle \citep{robinson2009travelling}. 
Identifying threats to migratory species, understanding species responses to change and developing effective conservation measures all require information on the movements of individual animals over multiple years, ideally for both historic and present-day populations. 
With the development of electronic tagging technology, studies of the distributions of animals have largely shifted from infrequent observations of many individuals with limited or no information about individual movement behavior, to frequent observations of a few individuals, with detailed information about individual movement \citep{holdo2013inferring}. 
Despite the tremendous advances made using direct telemetry devices, tags are expensive, with a relatively high failure rate \citep{bailey2009behavioural,best2015tag,mate2007evolution}. 
Data on individual-level, multi-annual movements remain scarce, especially for rare, wide-ranging and long-lived marine species such as baleen whales (Mysticeti) \citep{ryan2013stable,hall2005stable,bailey2009behavioural}.

An alternative technique for investigating animal movements retrospectively is to use intrinsic biochemical information such as stable isotope compositions \citep{west2006stable,busquets2017estimating,hobson2008tracking}. 
Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioral information that are often under-utilised \citep{lister2011natural}. 
The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
A large literature describes the use of stable isotope markers to link the composition of a tissue to a geographic source at a single point in time \citep{hobson2008tracking}. 
Relating the isotopic compositions of marine animal tissues to the likely location of tissue growth is complicated by a lack of knowledge of spatial variations in the isotopic composition of diet (the isotopic baseline) \citep{west2006stable,mcmahon2015millennial}. 
This uncertainty is compounded when multiple samples are taken through time in the same individual, as temporal variation in the isotopic baseline must also be considered explicitly. 
Consequently, relatively little attention has been given to the potential to infer individual movement histories by reconstructing time series of multiple discrete origins from sequentially sampled incrementally-grown tissues.

Here we introduce a new approach to infer location from the isotopic compositions of sequentially-sampled incrementally-grown tissues by drawing on newly developed models predicting spatio-temporal variation in phytoplankton stable carbon isotope composition at global and monthly resolution \citep{magozzi2017using} (Fig. S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1; Fig. S5).
We use our coupled models to simulate time series of carbon stable isotope compositions expected under different movement behaviours. 
We then compare simulated isotopic profiles with the measured profiles to identify potential movement histories most consistent with the measured isotopic profile.

We apply our method to blue whales (\textit{Balaenoptera musculus}), the largest animal to have ever lived. 
Mysticete whales are characterized by the development of baleen, keratinous structures in the upper jaw used to filter food items from seawater. 
Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert \citep{best1996stable,hobson1998stable}. 
Baleen plates therefore offer a continuous isotopic record of behavior typically reflecting multiple years of life of an individual whale. 
Baleen is worn away at the tips over time, so a baleen plate reflects the most recent years of life, and rarely records an individual's entire lifespan. 
Among the mysticete whales, blue whales are a particularly attractive target for model-based isotope movement work as they have a consistent, low trophic level diet and feed continuously through the year, likely driven by high energetic costs of maintaining extreme muscle mass \citep{goldbogen2015}.

Similar to other large balaenopterid whales, blue whales are generally assumed to conduct annual migrations between high and low latitudes \citep{huckgaete2018}. 
As blue whales feed throughout the year to accommodate energetic costs, migration routes and feeding areas for blue whales are thought to be shaped by the year-round location of highly productive regions \citep{branch2007}. 
Recent studies tracking movements of individual blue whales primarily in the northeast Pacific have demonstrated high levels of among-individual variation in movement histories including suspending migration for opportunistic feeding and skipped migration (i.e. year round residency in either summer or winter
areas) \citep{busquets2017estimating}. 
The northeast Pacific population of blue whales is perhaps the best studied in terms of individual-level movements with a recent history of deployment of over 177 satellite tags \citep{irvine2017quantifying, mate2007evolution}.
In the northeast Pacific, satellite tags show individuals occupying most of the exclusive economic zone (EEZ) and into the Gulf of Alaska \citep{irvine2017quantifying} with greatest overlapping areas in the southern half of the EEZ, along the continental slope and shelf region between the California Channel Islands and Cape Mendocino. 
Winter feeding is less well characterized, but winter feeding grounds are known
in the Gulf of California and in the eastern tropical Pacific, specifically the Costa Rica Dome where production is maintained by seasonal upwelling.
However, most satellite deployments on blue whales report movement data for periods of time under six months \citep{heide2001new, silva2013north, bailey2009behavioural, lesage2017foraging, irvine2017quantifying}, and are unable to show individual movements between summer and winter feeding or breeding areas. 
The longest period of continuous monitoring via satellite tagging of an individual blue whale that we are aware of in the northeast Pacific is 504 days \citep{irvine2017quantifying} and in the North Atlantic is 177 days \citep{lesage2017foraging}.

The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now around 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland \citep{pike2009note}. 
The small population size and oceanic habit makes blue whale movements extremely hard to study. 
A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggests that blue whales in the northeast Atlantic also track regions of high production throughout the year, at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde Islands \citep{baines2014upwellings}. 
Northward migration in the spring may occur along mid Atlantic corridors with peak sightings in the Azores around April-May \citep{silva2013north}. 
Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas \citep{pike2009note}. 
Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southerly migrating animals \citep{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
To our knowledge one photographic identification has matched summer and winter locations of a single north-east Atlantic blue whale with sightings in Iceland and Mauritania \citep{poster}.

@@

Here we infer individual level high-resolution movement records for seven blue whales for which baleen stable isotope data were available.
Isotope data for six whales stranding on the west coast of the USA or Mexico were reported in \cite{busquets2017estimating}, and we report new isotopic data from a whale stranding off the coast of Wexford Ireland in March 1891, and currently on display in the Natural History Museum, London (NHM).
The NHM whale ``Hope'' is a female estimated to be a young adult at least 15 years at death. 
Baleen from the NHM whale therefore yields information about migratory behaviors in North Atlantic blue whales around the peak of industrial whaling.

@@

Baleen was collected from the Natural History Museum, London (specimen NHMUK.1892.3.1.1). 
The baleen plate was cleaned with ethanol to remove surface contaminants such as skin/gum or other lipids that can influence isotopic signals. 
{\raise.17ex\hbox{$\scriptstyle\sim$}}1mg samples of keratin powder were then collected from the plate using a hand-held drill and grinding bit. 
97 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal (gingival) section that contains the most recent tissue. 
Baleen grows at a constant rate, so the samples are equally spaced through time \citep{best1996stable}. 
Carbon and nitrogen isotope analysis was performed simultaneously via continuous-flow isotope ratio mass spectrometry at the University of Southampton SEAPORT Stable Isotope Ratio Mass Spectrometry Laboratory (Southampton, UK), using a Vario Isotope select elemental analyser, coupled to an Isoprime 100 isotope mass spectrometer. 
Replicates using internal laboratory standards (L-glutamic acid (C), Glutamic acid (CT standard), acetanilide and protein standard OAS) were used for quality control and calibration. 
C:N ratios for samples ranged from 3.28\text{\textperthousand} to 3.72\text{\textperthousand}, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples \citep{hobson1998stable}. 
All data are available from the NHM Data Portal \citep{data-set} (https://doi.org/10.5519/0093278).

@@

Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates \citep{hobson1998stable,busquets2017estimating}. 
Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the youngest 70 - 18cm of the plate (behavioural phase two). 
We calculated isotopic periodicity within the NHM baleen sample using Fourier Transform analysis \citep{cardona2017temporal} (Supplementary Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated by \cite{busquets2017estimating}.  
Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891.

@@

Isotope-enabled biogeochemical ocean models \citep{magozzi2017using,schmittner2016complementary} were used to characterize
the isotopic composition of phytoplankton expected in different
potential foraging grounds (Fig. S3). 
Annual average \(\delta^{15}\)N POM (particulate organic matter) values were provided by C.J. Somes (\textit{pers.comm}) based on a 5\({}^{\circ}\) resolution biogeochemical model (Fig. S3). \(\delta^{13}\)C POM values were simulated at 1\({}^{\circ}\) and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model \citep{magozzi2017using,yool2013medusa}. 
Simulated \(\delta^{15}\)N POM values are relatively positive in the northeast Atlantic north of c. 60\({}^{\circ}\)N, and relatively negative in the central and southern North Atlantic. 
Annual average \(\delta^{13}\)C POM values largely vary with latitude, with more negative values in more northerly regions. 
In the central North Atlantic, \(\delta^{13}\)C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. S3). 
The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature \citep{magozzi2017using,laws1995dependence}. 
Thus temporal variations in \(\delta^{13}\)C POM values are superimposed on latitudinal gradients. 
The scale and nature of temporal variation in \(\delta^{13}\)C POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in \(\delta^{13}\)C POM values linked to strongly seasonal phytoplankton growth dynamics.
The model simulations used are forced with decadal climatological data from 2000-2010 \citep{magozzi2017using}. 
We do not draw inferences about location from absolute \(\delta^{13}\)C values and thus the reduction in oceanic \(\delta^{13}\)C POM values caused by release of fossil carbon into the atmosphere (Suess effect) does not influence inferences about location or movement. 
Ocean basin scale spatio-temporal differences in \(\delta^{13}\)C values largely reflect relative differences in sea surface temperatures between trophic, temperate and arctic waters, therefore we are confident that the \cite{magozzi2017using} model can be applied to interpret historic baleen isotope data.

We used \(\delta^{13}\)C POM values modeled at monthly resolution to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours. 
The stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on in the weeks prior to keratin growth. 
Assimilation of carbon into krill tissues will dampen the temporal variability seen in POM, effectively producing a temporal average over the timescale of isotopic turnover within krill. 
We estimate turnover to be complete between two and four months and therefore we resampled the \(\delta^{13}\)C POM values in each one degree cell to reflect an average of isotopic compositions in phytoplankton in the two months prior to the sampling date. 
Average values were weighted according to the proportional plankton biomass estimated for each month. 
Carbon isotope values are also likely to be fractionated during transfer from plankton to krill as \textsuperscript{12}C is preferentially lost through respiration. 
The degree of such trophic fractionation is unclear, however and we do not draw interpretations based on absolute \(\delta^{13}\)C values, rather on the relative \(\delta^{13}\)C values across the length of the baleen plate. 
We assume opportunistic ingestion of protein throughout the year \citep{baines2017autumn,silva2013north,visser2011timing,busquets2017estimating, lesage2017foraging, bailey2009behavioural, branch2007, huckgaete2018}.

@@

We simulated whale movements with the likelihood, direction, and extent of movement influenced by behavioral state, sea surface temperature, water depth, and phytoplankton concentration (as a proxy for zooplankton food availability).
Movement was coded as a set of probabilistic rules, informed by the literature on blue whale behavior (e.g. \citep{handbook}).
All terms were expressed as probability distributions, yielding multiple potential movement tracks.

In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and linear distance of movement are all influenced by the following. 
(i) Behavioural state (migrating north, migrating south, or foraging).
Northerly migrations were possible only in spring, and southerly migrations in autumn. 
Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton.
(ii) Sea surface temperature \citep{yool2013medusa} ($^{\circ}$C). 
When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters.  
(iii) Water depth (m; from the General Bathymetric Chart of the Oceans (GEBCO) global bathymetry dataset\citep{bathy}, extracted via the marmap package in R \citep{marmap}). 
Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
(iv) Phytoplankton concentration ($mmolNm^{-3}$, for combined diatom and non-diatom communities \citep{yool2013medusa}). 
This was included as a proxy for zooplankton food availability. 
Whales are more likely to move towards (or remain within) areas of high phytoplankton density, particularly during the foraging behavioural state.

@@

For the historic North Atlantic whale, we simulated the isotopicexpression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area \citep{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplemental Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
Measured tracks can only be replicated by combination of residency and seasonal migration with latitudinal migrations limited to the last four
years of simulations. 
For a full description of the model and its parameters see Supplementary Methods: Fig. S5 and Table S1. 
We simulated seven years of whale movements 1000 times and two years of residency 30 times for each residency hotspot (Fig. S4).
We simulated whale movements 1200 times, then excluded simulations where the virtual whale stranded before reaching the 3019 days of the baleen record, leaving 1049 simulations.

At each time point and location in the simulations, we extracted time averaged \(\delta^{13}\)C values from the spatio-temporal model \citep{magozzi2017using} resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figs X, \ref{fig2}, and S4).
We then compared the simulated stable isotope profiles (Figure \ref{fig3}) to the profile of the blue whale (Figure \ref{fig1}).

R code for all analyses is available from GitHub (https://github.com/nhcooper123/blue-whale-bes; \citep{github})

@@

Estimates of baleen growth rates for the six modern whales from the northeast Pacific were inferred from cyclical fluctuations in nitrogen isotope ratios and reported in \cite{busquets2017estimating}. 
We used these growth rates to reconstruct a potential timeline for baleen isotope records.

@@

In four of the six cases, measured baleen carbon isotope profiles could be replicated from model simulations with relatively high precision (Fig. \ref{pacific}
; best fit r\textsuperscript{2} values ranging from 0.43 to 0.74).
Whales D and F showed similar carbon isotope records and were best fit by the same movement simulation. 
We extracted monthly locations associated with the best fitting simulation models for each whale (Fig. \ref{pacific}).
Best fitting simulation models for whales D and F imply a relatively southerly distribution with few locations north of Baja, California and relatively limited seasonal latitudinal migration. 
The measured carbon isotope record for whale A was best fitted by simulations with a southerly distribution, but with an additional strong seasonal latitudinal migration, and with more northerly locations between April and October. 
Whale C was best fit by models with relatively northerly locations and strong seasonal latitudinal migration. 
We were unable to find convincing fits to baleen profiles for whales B and E suggesting that longer simulations with a greater variation in movements were needed. 
The inferred movement records are consistent with the known spatial distributions and relatively high level of individual variation in spatial distributions among blue whales in the Pacific northwest \citep{irvine2017quantifying, mate2007evolution}.
